# Determining Data Quality and Filtering

In this chapter, you'll learn:

- How to determine the quality of your data after a database search

- How to filter your data for downstream analysis

**Pre-requisites:**

- A basic understanding of R and RStudio

- A set of data searched using DIA-NN

- The following R packages installed from CRAN (recall, `install.packages("<package_name>"`): 

  - `tidyverse`
  
  - `BiocManager`
  
- The `limma` R package installed from Bioconductor (recall, `BiocManager::install("limma")`)

## Deciphering DIA-NN Search Result Files

After successfully completing a search, DIA-NN will create several files:

- **report.parquet**-  

- A variety of data matrices:

  - **report.pr_matrix.tsv**- Contains the abundances of all identified precursors with less than 5% FDR and a Q-value of <= 0.01. Each row is a unique precursor and each column is the abundance of that precursor in each sample. 
  
  - **report.pg_matrix.tsv**- Contains the abundances of all protein groups with less than 5% FDR and a Q-value of <= 0.01. Each row is a unique protein group and each column is the abundance of that protein group in each sample. 
    
  - **report.gg_matrix.tsv**- Contains the abundances of all gene groups with less than 5% FDR and a Q-value of <= 0.01. Each row is a unique gene group and each column is the abundance of that gene group in each sample. 
  
  - **report.unique_genes_matrix.tsv**- Contains the abundances of all unique genes (*i.e.*, only precursors that are uniquely identified to a gene are used for quantification) with less than 5% FDR and a Q-value of <= 0.01. Each row is a unique gene and each column is the abundance of that unique gene in each sample. 
  
  - **report.protein_description.tsv**- Contains a brief description of every protein group that was identified in the experiment. Each row will be a protein group and the columns will be the gene name, protein description of the first protein within the group, and the amino acid sequence of that protein.
  
  - **report.UniMod_modification-number_sites_90.tsv**- This is only created if scoring is set to "Peptidoform". Contains all the peptides with the specific modification where the localization was determined with >= 90% confidence. Each row is a unique modified peptided=s and each column is the abundance of that peptide in each sample.
  
  - **report.UniMod_modification-number_sites_99.tsv**- This is only created if scoring is set to "Peptidoform". Contains all the peptides with the specific modification where the localization was determined with >= 99% confidence. Each row is a unique modified peptides and each column is the abundance of that peptide in each sample.

- **report.stats.tsv**- Contains basic performance metrics for each run such as the number of protein and precursor IDs. Each row is a different sample and each column is a different metric. 

- **report.log.txt**- The text output from the DIA-NN GUI dialog box. Can be useful to recall search settings. 

- **report-lib.parquet**- This is only created if you check the box "create spectral library" (automatically enabled if you enable "MBR"). Contains the spectral library created from the first-pass search results. 

- **report_runs.pdf**- Contains summary plots for each run used in the search.

- **report_trends.pdf**- Contains summary plots for basic metrics to compare all runs used in the search.

- If you used match-between-runs (MBR), all the above listed will be the results from the MBR search. There will also be a second group of files all with the same names as the files above but have "-first-pass" after "report". 

## Import DIA-NN Search Results

First we need to load the libraries we'll need:

```{r}
library(arrow)
library(tidyverse)
```

Next, let's load in all the files we're going to be using

```{r, cache=TRUE}
diann_stats <- read_tsv('report.stats.tsv')
diann_results <- read_parquet('report.parquet')
protein_names <- read_tsv('report.protein_description.tsv')
```

The data we're going to be using for this guide was generated by Maria Ahmed, a student in the Ghaemmaghami lab. These are a series of *E. coli* gene knockouts of the **Methionine Sulfoxide Reductase* gene family responsible for reducing oxidized methionines and are from the Keio *E. coli* Knockout Collection. 

## Evaluate DIA-NN Stats

Before we go too far, let's look at our stats report file to see if anything the experiment looks "off".

To look at the data in this file, use `View()`.

```{r, eval=FALSE}
View(diann_stats)
```

```{r, echo=FALSE}
diann_stats
```

What you'll see is each run as it's own row and then 16 different "measurements" of each run. The columns you are going to use most of the time are:

- **Precursors.Identified**- The number of *precursors* identified in that run. It is important to note that a precursor is simply a charged peptide, so a single peptide sequence can have multiple precursors (*i.e.*, if peptide *x* was both doubly and triply charges, that would count as 2 precursor IDs)
  

- **Proteins.Identified**- The number of *proteins* identified in that run. 

- **Average.Missed.Tryptic.Cleavages**- The average number of missed cleavages by trypsin in that run. A lower number indicates a more efficient digestion.

- Some typical numbers for various experiments might be:
  
  - Global experiment of mammalian cell lysate: 
    
    - 90,000-140,000 Precursor IDs
    
    - 9,000-10,000 Protein IDs
    
    - 5-15% Missed Cleavages
    
  - Global experiment of *E. coli* lysate: 
  
    - 35,000-45,000 Precursor IDs
    
    - 3,000-3,300 Protein IDs
    
    - 10-20% Missed Cleavages
    
  - Phospho experiment (varies greatly on type of material and amount enriched; assuming 100ug of cultured cells): 
  
    - 30,000-40,000 Precursor IDs
    
    - 4,000-5,500 Protein IDs
    
    - 40-60% Missed Cleavages 
    
  - Single-cell experiment:
      
    - 2,000-10,000 Precursor IDs (varies greatly depending on cell type and size)
    
    - 500-2,500 Protein IDs (varies greatly depending on cell type and size)
    
    - 10-20% Missed Cleavages
    
Since the stats report has a lot of information that isn't necessarily important for a majority of users, let's look at the columns we're interesting in using `dplyr::select()`. 

Additionally, the file name unfortunately includes the entire file path of the raw file, making it harder to see and understand what each file is. We're also going to *pipe* in a `dplyr::mutate()` to change the `File.Name` column to remove the common text in all the rows using the `gsub` function. This function looks for all strings that match the one entered in and then replaces those strings with one you tell it to (in this case it will be nothing or `""`). And within this `gsub()`, you'll also be seeing out first example of a **regular expression** or **regex**. **Regex** commands can be useful when working with lnong strings following a standardized "formula". In this scenario, all files use the following string `Y:\\9- Teaching Materials and Presentations\\Proteomics Summer Course\\2026\\Datasets\\E Coli Ahmed\\Raw Files\\Ahmed_` followed by the sample name. Instead of type out the first part, which is the same in every sample, we're going to type `Y.*Ahmed_`, which simply means "Any string that starts with *Y* followed by *any character* (`.`) for an *unlimited number of times* (`*`), until we reach the string "Ahmed_", at which point, we stop. And then that entire string will be replaced with `""`, which is nothing. 
(
We are also going to remove the ending string from each run since those are also the same for every one; "_25-023.raw" (the work order number used by the MSRL assigned to each project, followed by the file type of that run).

```{r}
diann_stats %>% 
  # We only want some specific columns
  select(File.Name, # Name of the run
         Precursors.Identified, # The number of Precursor IDs
         Proteins.Identified, # The number of Protein IDs
         Average.Missed.Tryptic.Cleavages) %>% # The average number of missed cleavages in that run
  # We are going to the common sample monikers the MSRL uses. Specifically, each run starts with "<Researcher name>_" and ends with "_<Work order number>". For the dataw e will be using in this tutorial, the researcher name is "Ahmed" and the work order number is "25-023".
  # We are going to combine the 'mutate()' dplyr function with the 'sub()' base R function to change the contents of the "File.Name" column. Specifically, we are going to find all times either "Amhed_" or "_25-023" appear in the the column "File.Name" and replace those strings with "" (ie, nothing)
  mutate(File.Name = sub('Y.*Ahmed_', #
                         '',
                         File.Name),
         File.Name = sub('_25-023.raw',
                         '',
                         File.Name))
```

Just by looking at these numbers, it appears as if the data are all relatively consistent (with the exception of 1 file), at least from a very wide lens. What this is not helpful with though, is determining if there are any micro-trends within the samples to indicate how well the experiment performed. 

But we can be confident the runs all performed well compared to each other and that there are no major outliers or nothing concerning at first glance. 

## Evaluate DIA-NN Results

Let's look at the data in this file. To do so, we'll use `head()` to look at the first five rows of the data frame.

```{r}
head(diann_results)
```

### Explaining the DIA-NN Report

Each row is a **precursor** ID (*i.e.*, a charged peptide...remember you can have multiple precursors for a peptide) and each row is data specific to that precursor. A brief description of some of the important rows are below:

- **Run**: the run name the precursor was identified from

- **Precursor.Id**: the amino acid sequence followed by charge state of the precursor. Note that if the precursor is modified, the UniMod number will be here as well

- **Modified.Sequence**: the amino acid sequence of the precursor. Note that is the that if the precursor is modified, the UniMod number will be here as well

  - *Note*: it is possible to have multiple Precursor.Ids for the same Modified.Sequence (*e.g.*, the modified peptide was identified in both the +2 and +3 form)

- **Stripped.Sequence**: the amino acid sequence of the precursor. Note that if the precursor is modified, the UniMod number will **NOT** be here. 

  - *Note*: it is possible to have multiple Modified.Sequences for the same Stripped.Sequence (*e.g.*, the peptide was identified in both the modified and un-modified form)
  
- **Precursor.Charge**: the charge state of the precursor

- **Proteotypic**: the precursor sequence matches to a single protein in the FASTA file. (1 = TRUE, 0 = FALSE)

- **Precursor.mz**: the *m/z* of the identified precursor

- **Protein.Ids**: The UniProt accession number(s) of all possible proteins the precursor matched to (*i.e.*, every protein this precursor could have come from)

- **Protein.Group**: DIA-NN's best attempt at resolving which minimal set of proteins can explain the identified precursor (*i.e.*, what DIA-NN **thinks** the precursor most likely came from)

- **Genes**: The name of the gene that produces the "Protein.Group"

- **RT**: The retention time, in minutes, of the apex for the precursor

- **IM**: The ion mobility (if the instrument has it) of the precursor

- **Precursor.Quantity**: The un-normalized abundance of the precursor

- **Precursor.Normalised**: The normalized abundance of the precursor. Note that if normalization is disabled. the column ios simply the same value as "Precursor.Quantity"

- **Quantity.Quality**: The quality of the precursor quantification (higher value = higher confidence)

- **RT.Start**: The retention time, in minutes, of the first scan the precursor was detected

- **RT.Stop**: The retention time, in minutes, of the last scan the precursor was detected

- **FWHM**: The "Full-width, half-maximum" of the precursor, in minutes. This gives a measurement for how wide the precursor extracted ion chromatogram (XIC)

- **PG.MaxLFQ**: The abundance of the entire protein group that precursor matched to. This is calculated by using "MaxLFQ", which tries to use all the possible precursor abundances across all the runs to find the most representative abundance that describes that protein group's abundance for each sample. 

  - *Note*: this value will be the same for all precursors that match to the same protein group within each sample
  
- **PG.MaxLFQ.Quality**: The quality of the MaxLFQ quantification (high value = higher confidence)

- **Q.Value**: The confidence that the precursor identification within each run was a true identification (lower value = higher confidence)

- **PEP**: "Posterior Error Probability" (lower value = higher confidence)

- **Global.Q.Value**: The confidence that the precursor identification across all runs in the experiment was a true identification (lower value = higher confidence)

- **Lib.Q.Value**: The confidence that the precursor identification was matched to a precursor from the spectral library (lower value = higher confidence)
  
  - *Note*: this only populate if MBR was enabled

- **PG.Q.Value**: The confidence that the protein group identification within each run was a true identification (lower value = higher confidence)

- **Global.PG.Q.Value**: The confidence that the protein group identification across all runs in the experiment was a true identification (lower value = higher confidence)

- **Lib.PG.Q.Value**: The confidence that the protein group identification was matched to a protein group from the spectral library (lower value = higher confidence)
  
  - *Note*: this only populate if MBR was enabled

- **Best.Fr.Mz**: The *m/z* of the best fragment ion for this precursor. This isn't always used, but will be necessary if you wish to use `MSStats`

### Evaluating the Quality of the Runs

Using some of the columns mentioned above, we are going to look at the quality of each of the precursors, and filter to only retain high quality IDs, and then re-check those same metrics.

But first, we're going to modify the `diann_results` data frame slightly:

- We will rename each `Run` so that we remove any common strings ("Ahmed_" and "_25-023"). This will make visualization slightly easier

- We will remove any unwanted columns so that we only have the ones mentioned above. The makes the data frame considerably smaller, making each subsequent step just a bit faster

- We will also be adding an extra column denoting the group of each sample. This will make it slightly easier to comprehend the graphs later on

```{r}
diann_results <- diann_results %>% 
  select(c(Run, Precursor.Id:Precursor.Charge, Proteotypic:Protein.Group,
           Genes:RT, Precursor.Quantity, Precursor.Normalised, Quantity.Quality, 
           RT.Stop:FWHM, PG.MaxLFQ, PG.MaxLFQ.Quality, Q.Value:Lib.Q.Value,
           PG.Q.Value, Global.PG.Q.Value, Lib.PG.Q.Value)) %>% 
  mutate(Run = gsub('Ahmed_', '', Run),
         Run = gsub('_25-023', '', Run)) %>% 
  # We are going to add "metadata" to our runs, which includes information such as the group that sample belongs to, the replicate, etc. We can make this beforehand, or simply make a new tibble that contains all the information, which we're doing here. So we will be making a new tibble where one column is all of the run names in our experiment ("Runs") and then another column that has the genotype of that samples ("group")
  left_join(y = tibble(Run = c( 'bisC1', 'bisC2', 'bisC3', 
                                'msrA1', 'msrA2', 'msrA3',
                                'msrB1', 'msrB2', 'msrB3',
                                'msrC1', 'msrC2', 'msrC3', 
                                'WT1', 'WT2', 'WT3'),
                       group = c(rep('bisC', 3),
                                 rep('msrA', 3),
                                 rep('msrB', 3),
                                 rep('msrC', 3),
                                 rep('WT', 3))),
            by = 'Run')
```

#### Proteome Coverage

Now let's look to see how many precursors, modified peptides, and protein groups were identified in each experiment. Ideally, they should be pretty similar, or at least each biological group should be similar among each other. 

First, we will look at the number of precursors identified in each sample. While it would be nice to have all samples have a similar number, sometimes biology gets in the way. At the very least, we should expect each replicate from the same group to have a similar number of IDs. We'll do this by looking at three different ID "levels": precursors, peptides, and proteins.

##### Precursor IDs {-}

```{r}
diann_results %>% 
  # We want only the columns "Run", "group", and "Precursor.Id". More specifically, only rows where all three of these columns are unique
  distinct(Run,
           group,
           Precursor.Id) %>% 
  group_by(Run,
           group) %>% 
  summarise(n_precursor_ids = n(), #counts the number of precursor IDs in each group (run and group)
            .groups = 'drop') %>% 
  # We want to plot the number of unique precursors in each sample, grouping them by genotype
  ggplot(mapping = aes(x = Run, # x-axis will be the run
                       y = n_precursor_ids, # y-axis will be the number of precursor IDs
                       fill = group)) + # we want each genotype to be a different color
  geom_col() + # Draw a bar graph
  # Add labels to the plot
  labs(x = 'Sample', # x-axis title
       y ='Number of Precursor IDs', # y-axis title
       title = 'Number of Precursor IDs per Run') #Title of the plot
```

##### Peptide IDs {-}

```{r}
diann_results %>% 
  distinct(Run,
           group,
           Modified.Sequence) %>% 
  group_by(Run,
           group) %>% 
  summarise(n_modified_peptides = n(),
            .groups = 'drop') %>% 
  ggplot(mapping = aes(x = Run,
                       y = n_modified_peptides,
                       fill = group)) +
  geom_col() +
  labs(x = 'Sample',
       y ='Number of Modified Peptide IDs',
       title = 'Number of Modified Peptide IDs per Run')
```

##### Protein IDs {-}

```{r}
diann_results %>% 
  distinct(Run,
           group,
           Stripped.Sequence) %>% 
  group_by(Run,
           group) %>% 
  summarise(n_peptides = n(),
            .groups = 'drop') %>% 
  ggplot(mapping = aes(x = Run,
                       y = n_peptides,
                       fill = group)) +
  geom_col() +
  labs(x = 'Sample',
       y ='Number of Peptide IDs',
       title = 'Number of Unique Peptide IDs per Run')

diann_results %>% 
  distinct(Run,
           group,
           Protein.Group) %>% 
  group_by(Run,
           group) %>% 
  summarise(n_protein_groups = n(),
            .groups = 'drop') %>% 
  ggplot(mapping = aes(x = Run,
                       y = n_protein_groups,
                       fill = group)) +
  geom_col() +
  labs(x = 'Sample',
       y ='Number of Protein Groups',
       title = 'Number of Protein Group IDs per Run')
```

This series of plots show us a few things:

1. The number of IDs decreases as we go up in "biological level" (*i.e.*, precursor -> modified peptide -> stripped peptide -> protein). This inherently makes sense since multiple precursors can be produced from a single modified peptide, if there is a variable modification site, a peptide can exist as both modified and unmodified, and lastly multiple peptides can be produced from a single protein

2. There seems to be one run, msrC2, that is an outlier in terms of IDs as compared to the rest of the runs, especially the rest of the runs from the group. 

However, based just off the total number of IDs, we can't tell if we should remove this run or not from our analysis. Let's look at some other metrics, specifically:

- Charge state distribution

- Intensity distribution

- CV distribution at each "biological level" (precursor, peptide, protein group)

#### Charge State Distribution

```{r}
diann_results %>% 
  select(Run, 
         group,
         Precursor.Id,
         Precursor.Charge) %>% 
  count(Run,
        group,
        Precursor.Charge) %>% 
  group_by(Run, group) %>% 
  mutate(total_precursors = sum(n)) %>% 
  group_by(Run, group, Precursor.Charge) %>% 
  reframe(charge_per = n/total_precursors * 100) %>% 
  mutate(Precursor.Charge := fct_inorder(factor(Precursor.Charge)),
         Run := factor(Run,
                       levels = unique(str_sort(Run, numeric = T)))) %>% 
  ggplot(aes(x = Run,
             y = charge_per,
             fill = Precursor.Charge)) +
  geom_col(color = 'black',
           linewidth = 0.75) +
  facet_grid(cols = vars(group),
             scales = 'free') +
  labs(y = 'Percentage of Precursors',
       title = 'Percentage of Precursors at Specific Charge Among All IDs') +
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
```

All runs follow a similar pattern of charge distribution, indicating ionization was consistent across the experiment

#### Intensity Distribution {-}

##### Un-normalized Intensity {-}

```{r}
diann_results %>% 
  select(Run, 
         group,
         Precursor.Id,
         Precursor.Quantity) %>%
  filter(Precursor.Quantity > 0) %>% 
  mutate(Precursor.Quantity = log2(Precursor.Quantity)) %>% 
  ggplot(aes(x = Precursor.Quantity,
             color = Run)) +
  geom_freqpoly(linewidth = 0.75,
                bins = 25) +
  facet_grid(cols = vars(group),
             scales = 'free') +
  labs(x = 'Un-normalized Precursor Abundance (log2)',
       title = 'Distribution of Un-normalized Precursor Abundances') 
```

It would appear that the `msrC2` sample might have had lower coverage due to a decrease in the total amount of peptide loaded onto the HPLC column, which would in turn lead to a decrease in overall abundances. Luckily, normalization can sometimes make up for differences like these. Let's look to see if DIA-NN normalization helped. 

##### Normalized Intensity

```{r}
diann_results %>% 
  select(Run, 
         group,
         Precursor.Id,
         Precursor.Normalised) %>%
  filter(Precursor.Normalised > 0) %>% 
  mutate(Precursor.Normalised = log2(Precursor.Normalised)) %>% 
  ggplot(aes(x = Precursor.Normalised,
             color = Run)) +
  geom_freqpoly(linewidth = 0.75,
                bins = 25) +
  facet_grid(cols = vars(group),
             scales = 'free') +
  labs(x = 'Normalized Precursor Abundance (log2)',
       title = 'Distribution of Normalized Precursor Abundances') 
```

While the left edge of the curve is narrower, `msrC2` now mostly agrees with the rest of the group. This would imply that by using the normalized precursor quantities, we shouldn't see too much effect from the decreased material loaded in `msrC2` on the downstream analyses. Just to be sure, let's check the protein abudances, which is calculated using the normalized precursor abundances.

```{r}
diann_results %>% 
  distinct(Run, 
           group,
           Protein.Group,
           PG.MaxLFQ) %>%
  filter(PG.MaxLFQ > 0) %>% 
  mutate(PG.MaxLFQ = log2(PG.MaxLFQ)) %>% 
  ggplot(aes(x = PG.MaxLFQ,
             color = Run)) +
  geom_freqpoly(linewidth = 0.75,
                bins = 25) +
  facet_grid(cols = vars(group),
             scales = 'free') +
  labs(x = 'Protein Abundance (log2)',
       title = 'Distribution of Protein Abundances') 
```

Once "wrapping up" to the protein level, any differences in abundances is now even smaller. Therefore, we can be a little more confident that we can use all our data in downstream analyses. But just to be sure, we're going to look at one more metric. 

#### CV Distribution {-}

We are also going to measure the coefficient of variation (CV, standard deviation divided by the mean) across each "biological level" to see if there are any groups what have inconsistent quantification across all reps of that group. 

We will also be measuring the CV across all 15 samples, regardless of biological group. This is a good sanity check to always perform; if your experiment was properly designed, the CV between replicates within the same group should be lower than the CV between all the samples.

##### Precursor Level {-}

```{r}
diann_results %>% 
  distinct(Precursor.Id,
           Run,
           group,
           Precursor.Normalised) %>% 
  filter(Precursor.Normalised > 0) %>% 
  group_by(Precursor.Id) %>% 
  mutate(cv_combined = sd(Precursor.Normalised,
                          na.rm = T)/mean(Precursor.Normalised,
                                          na.rm = T) * 100) %>% 
  ungroup() %>% 
  group_by(group,
           Precursor.Id) %>% 
  mutate(cv = sd(Precursor.Normalised,
                 na.rm = T)/mean(Precursor.Normalised,
                                 na.rm = T) * 100) %>% 
  ungroup() %>% 
  distinct(group,
           Precursor.Normalised,
           cv_combined,
           cv) %>% 
  drop_na() %>% 
  pivot_longer(cols = starts_with('cv'),
               names_to = 'type',
               values_to = 'cv') %>% 
  mutate(type = ifelse(type == 'cv',
                       group,
                       'All Groups Combined')) %>% 
  mutate(type = fct_relevel(as.factor(type),
                            'All Groups Combined')) %>% 
  ggplot(aes(x = type,
             y = cv)) +
  geom_violin(aes(fill = type),
             na.rm = T) +
  geom_boxplot(na.rm = T,
               width = 0.1) +
  labs(title = 'Distribution of Coefficient of Variation: Precursor-level',
       x = '',
       y = 'Coefficient of Variation (%)',
       fill = 'Group') +
  scale_y_continuous(limits = c(0, 200))
```

For all groups, the CV of the normalized precursors are all below the intra-group CV, including the msrC group, which had the one outlier. However, based on our earlier graphs, we should expect the CV of that group to decrease as we "go up" in biological level to the protein level.

##### Protein Level {-}

```{r}
diann_results %>% 
  distinct(Protein.Group,
           Run,
           group,
           PG.MaxLFQ) %>% 
  filter(PG.MaxLFQ > 0) %>% 
  group_by(Protein.Group) %>% 
  mutate(cv_combined = sd(PG.MaxLFQ,
                          na.rm = T)/mean(PG.MaxLFQ,
                                          na.rm = T) * 100) %>% 
  ungroup() %>% 
  group_by(group,
           Protein.Group) %>% 
  mutate(cv = sd(PG.MaxLFQ,
                 na.rm = T)/mean(PG.MaxLFQ,
                                 na.rm = T) * 100) %>% 
  ungroup() %>% 
  distinct(group,
           PG.MaxLFQ,
           cv_combined,
           cv) %>% 
  drop_na() %>% 
  pivot_longer(cols = starts_with('cv'),
               names_to = 'type',
               values_to = 'cv') %>% 
  mutate(type = ifelse(type == 'cv',
                       group,
                       'All Groups Combined')) %>% 
  mutate(type = fct_relevel(as.factor(type),
                            'All Groups Combined')) %>% 
  ggplot(aes(x = type,
             y = cv)) +
  geom_violin(aes(fill = type),
              na.rm = T) +
  geom_boxplot(na.rm = T, 
               width = 0.1) +
  labs(title = 'Distribution of Coefficient of Variation: Protein-level',
       x = '',
       y = 'Coefficient of Variation (%)',
       fill = 'Group') +
  scale_y_continuous(limits = c(0, 200))
```

While the CV of the msrC group did go down, so did the intra-group CV. So much so that it is now lower than the msrC CV. This means that the msrC group has too much variability to reliably compare to other groups.

However, we know that one run in particular is causing this variability. So what would have if we were to remove that run (msrC2)?

```{r}
diann_results %>% 
  distinct(Protein.Group,
           Run,
           group,
           PG.MaxLFQ) %>% 
  filter(Run != 'msrC2') %>% 
  filter(PG.MaxLFQ > 0) %>% 
  group_by(Protein.Group) %>% 
  mutate(cv_combined = sd(PG.MaxLFQ,
                          na.rm = T)/mean(PG.MaxLFQ,
                                          na.rm = T) * 100) %>% 
  ungroup() %>% 
  group_by(group,
           Protein.Group) %>% 
  mutate(cv = sd(PG.MaxLFQ,
                 na.rm = T)/mean(PG.MaxLFQ,
                                 na.rm = T) * 100) %>% 
  ungroup() %>% 
  distinct(group,
           PG.MaxLFQ,
           cv_combined,
           cv) %>% 
  drop_na() %>% 
  pivot_longer(cols = starts_with('cv'),
               names_to = 'type',
               values_to = 'cv') %>% 
  mutate(type = ifelse(type == 'cv',
                       group,
                       'All Groups Combined')) %>% 
  mutate(type = fct_relevel(as.factor(type),
                            'All Groups Combined')) %>% 
  ggplot(aes(x = type,
             y = cv)) +
  geom_violin(aes(fill = type),
              na.rm = T) +
  geom_boxplot(na.rm = T, 
               width = 0.1) +
  labs(title = 'Distribution of Coefficient of Variation: Protein-level',
       x = '',
       y = 'Coefficient of Variation (%)',
       fill = 'Group') +
  scale_y_continuous(limits = c(0, 200))
```

Our msrC group now has really low CVs, similar to the other runs. Therefore in downstream analyses, we have to remember to remove that run before doing anything else.

## Filtering DIA-NN Results

Now that we have a good grasp on how our data looks, we're going to remove any low-quality and contaminant precursors from our dataset before we move forward with any analyses. Additionally, since we know that `msrC2` is an outlier, we are going to move any IDs from that sample as well.

- We will be removing "low quality" IDs through use of the various `Q.Value` columns and quantification `Quality` columns. 

- We will be removing contaminant IDs through using the "cRAPome" database that we included in our spectral library
  
  - Recall from the previous chapter that using the cRAPome will add common mass spec contaminants to our spectral library. Anything that matches to one of those proteins will then have "cRAP" in any of: `Protein.Ids`, `Protein.Group`, `Protein.Ids`, and/or `Genes`
  
```{r}
diann_filter <- diann_results %>% 
  filter(Run != 'msrC2') %>% #remove the outlier run
  #remove the poor quality IDs
  filter(Q.Value <= 0.01 &
           Lib.Q.Value <= 0.01 &
           PG.Q.Value <= 0.05 &
           Lib.PG.Q.Value <= 0.05 &
           Quantity.Quality > 0 &
           PG.MaxLFQ.Quality > 0) %>% 
  #remove the contaminant IDs
  filter(!grepl('cRAP', Protein.Ids) &
         !grepl('cRAP', Protein.Group) &
         !grepl('cRAP', Protein.Ids) &
         !grepl('cRAP', Genes)) %>% 
  select(c(Run, group, Modified.Sequence, Protein.Group, Genes, PG.MaxLFQ))
```

To give an idea as to the number of IDs we removed:

```{r}
full_join(x = diann_results %>% 
            group_by(Run) %>% 
            mutate(n_precursors_preFilter = n()) %>% 
            distinct(Run, n_precursors_preFilter),
          y = diann_results %>% 
            distinct(Run, Protein.Group) %>% 
            group_by(Run) %>% 
            mutate(n_proteins_preFilter = n()) %>% 
            distinct(Run, n_proteins_preFilter),
          by = 'Run') %>% 
  full_join(y = diann_filter %>% 
            group_by(Run) %>% 
            mutate(n_precursors_postFilter = n()) %>% 
            distinct(Run, n_precursors_postFilter),
            by = 'Run') %>% 
  full_join(y = diann_filter %>% 
              distinct(Run, Protein.Group) %>% 
              group_by(Run) %>% 
              mutate(n_proteins_postFilter = n()) %>% 
              distinct(Run, n_proteins_postFilter),
            by = 'Run') %>% 
  arrange(Run)
```

## Prepare Data for Differential Expression Analysis (t-test)

There are just a few more things we need to do before our data is ready for statistical analysis. More specifically, we need to:

1. Filter data based off total number of peptides and completeness

2. Replace missing values through imputation

### Filter Based off Number of Peptides and Missing Values

When dealing with biological data, especially when dealing with gene knockouts, it would not be unexpected for some proteins to not be detected at all in some groups compared to others. This could be a result of a combination of two things:

1. The protein is either down-regulated or not even expressed in one group

2. The level of the protein is beneath the limit of detection of our analytical method (in this case, the mass spectrometer)

The main issue that arises from this is that some proteins will be detected in some samples/groups and not others. The samples the protein was not detected in will have a value like `0` of `NA` denoting that there is no value present for that protein in that sample. This is called a **missing value**. R does not handle `NA` very well (or any programming language for that matter), so when we try to perform any kind of statistical analysis on groups where `NA` is present, it either will not work or will look "funky". There are several ways to circumvent this program, including:

- Removing any proteins with a missing value

- Replacing missing values with a constant number

- **Imputing** the missing value using some mathematical formula

The last method, imputation is usually how we go ahead with handling missing values. Specifically, a method where a random, low-value number is assigned to a missing value (at the protein level). More specifically:

1. The protein abundances are converted to log2 (to make the data normally distributed)

2. A standard curve is made from the log2 protein abundances

3. That standard curve is shifted by a standard deviation we determine (we default to 1.8 standard deviations)

4. The new, shifted curve is "shrunk" to a width of the original curve that we determine (we default to 0.3 the width of the original curve)

5. A random number from this new, shifted and shrunken curve is assigned to each missing value

A key assumption to this method is that any missing values correspond to low-abundance proteins that fall below the limit of detection of our mass spectrometer.

However, before we impute, we perform one more quality check to make sure what we're imputing isn't artificially creating biological phenomena. Specifically, we set thresholds for each protein so that each protein:

1. Has multiple peptides that map to that protein (no "one hit wonders")

2. There are multiple replicates within at least one group with multiple peptides

```{r}
diann_peptide_counts <- diann_filter %>% 
  group_by(Run, Protein.Group, Genes) %>% 
  summarise(n_peptides = n(),
            .groups = 'drop') %>% 
  left_join(y = diann_filter %>% 
              distinct(Run, group, Protein.Group, Genes, PG.MaxLFQ),
            by = c('Run', 'Protein.Group', 'Genes'))

diann_peptide_filter_list <- diann_peptide_counts %>% 
  drop_na() %>% 
  select(-PG.MaxLFQ) %>% 
  filter(n_peptides >= 2) %>% 
  group_by(group, Protein.Group, Genes) %>% 
  summarise(n_reps = n(),
            .groups = 'drop') %>% 
  filter(n_reps >= 2) %>% 
  select(-n_reps)

diann_peptide_filtered_data <- inner_join(x = diann_peptide_counts,
                                          y = diann_peptide_filter_list,
                                          by = c('group', 'Protein.Group', 'Genes'))

diann_peptide_filtered_data
```

We now have a data frame with a new column containing the number of peptides that were identified for each protein in each sample. And was filtered so that every protein was quantified using 2 or more peptides. AND has at least one group with at least two replicates with that protein present. So now we are a little more confident in our dataset that the proteins that are still present were at least quantified in at least one of our groups. 

The last thing we need to do before imputing missing values is to make our data "wide", which just means that instead of each run being on its own row, each row is now a single protein and each column is the protein abundance in that sample. At this point, we will now see `NA` if there is a missing value.

```{r}
diann_peptide_filtered_wide <- diann_peptide_filtered_data %>% 
  select(-c(n_peptides, group)) %>% 
  mutate(PG.MaxLFQ = log2(PG.MaxLFQ)) %>% 
  pivot_wider(names_from = 'Run', 
              values_from = 'PG.MaxLFQ')

diann_peptide_filtered_wide
```

### Impute Missing Values

With this wide data frame, we will now impute missing values. 

We have an internally created function to perform this imputation, it is as follows:

```{r, echo=FALSE}
impute_normal <- function(df,
                          width     = 0.3,
                          downshift = 1.8,
                          seed      = 100) {
  
  if (!is.matrix(df))
    df <- as.matrix(df)
  mx <- max(df, 
            na.rm = T)
  mn <- min(df, 
            na.rm = T)
  if (mx - mn > 25)
    warning("Please make sure the values are log-transformed")
  
  set.seed(seed)
  df <- apply(df,
              2,
              function(temp) {
                temp[!is.finite(temp)] <- NA
                temp_sd <- stats::sd(temp, 
                                     na.rm = T)
                temp_mean <- mean(temp, 
                                  na.rm = T)
                shrinked_sd <- width * temp_sd
                downshifted_mean <- temp_mean - downshift * temp_sd
                n_missing <- sum(is.na(temp))
                temp[is.na(temp)] <- stats::rnorm(n_missing,
                                                  mean = downshifted_mean,
                                                  sd   = shrinked_sd)
                temp
              })
  return(df)
  
}
```

```{r}
imputed_values <- impute_normal(df = diann_peptide_filtered_wide %>%  
                                  select(-Genes) %>% 
                                  column_to_rownames(var = 'Protein.Group'),
                                width = 0.3,
                                downshift = 1.8) 

head(imputed_values)
```

Now there should be no missing values present in the data. The next thing we need to do is make this data back into a format more convenient to work with (*i.e.*, a tibble) and add back in particular columns (*e.g.* `Protein.Group` and `Genes`)

```{r}
imputed_data_wide_df <- imputed_values %>% 
  as.data.frame() %>% 
  rownames_to_column(var = 'Protein.Group') %>%
  bind_cols(y = diann_peptide_filtered_wide %>% 
              select(Genes)) %>% 
  relocate(Genes, .after = 'Protein.Group')

imputed_data_df <- imputed_data_wide_df %>% 
  pivot_longer(cols = 3:last_col(),
               names_to = 'Run',
               values_to = 'PG.MaxLFQ') %>% 
  left_join(diann_peptide_filtered_data %>% 
              select(-c(PG.MaxLFQ, group)),
            by = c('Run', 'Protein.Group', 'Genes')) %>% 
  left_join(diann_peptide_filtered_data %>% 
              distinct(Run, group),
            by = 'Run') %>% 
  relocate(Run, .before = 'Protein.Group') %>% 
  relocate(group, .after = 'Run')

imputed_data_df
```

Now we have a long data frame that contains protein abundances with no missing values along with the number of peptide IDs (if applicable) for that protein. Now we can perform differential expression analysis using a t-test from `limma`, which we will be doing in the next chapter. 